---
title: "NYCHVS Data Analysis"
author: Analyzing New York City housing data over the past three decades for the *UVA Cavalytics Spring Data Competition* and annual *American Statistical Association's* (ASA) <a href="https://www1.nyc.gov/site/hpd/about/nychvs-asa-data-challenge-expo.page?fbclid=IwAR0fN8TUt2UfJc3qzglI-MtMoG0WW_mky-b5Njv5yd2MdsO4Z4lp0WKiOxo#"><target = "blank">Data Challenge Expo</a>
output:
  github_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Research Question

The majority of housing units in NYC are renter-occupied according to this <a href="https://www.census.gov/programs-surveys/ahs/visualizations/metrobriefs-2013.html"><target = "blank">American Housing Survey</a> -- slightly over half of the city, as they found --  and so I decided to focus on renters for my research. Specifically, I am interested in how spending behavior on the essential expense of shelter has changed since the early 90's. **Are renters of NYC spending more or less relative to what they make in income?** What about those with mortgages or some sort of loan? Is there a change in spending behavior and if so, is the change consistent among renters and mortgagers, or among different demographics?

## Hypothesis

Due to record high debt levels in America, the expectation is that spending behavior on housing somewhat reflects that. That said, I foresee that residents of NYC are spending more than is comfortable to spend or more than what was reasonable for them to spend in years past. 

## Data Cleaning

*Feel free to skip past this to the analysis portion*

```{r}
library(tidyverse)

csv91 <- "/Users/malcolm_mashig/Downloads/NYCHVS 1991 Occupied File for ASA Challenge_CSV.csv"
csv93 <- "/Users/malcolm_mashig/Downloads/NYCHVS 1993 Occupied File for ASA Challenge_CSV.csv"
csv96 <- "/Users/malcolm_mashig/Downloads/NYCHVS 1996 Occupied File for ASA Challenge_CSV.csv"
csv99 <- "/Users/malcolm_mashig/Downloads/NYCHVS 1999 Occupied File for ASA Challenge_CSV.csv"
csv02 <- "/Users/malcolm_mashig/Downloads/NYCHVS 2002 Occupied File for ASA Challenge_CSV.csv"
csv05 <- "/Users/malcolm_mashig/Downloads/NYCHVS 2005 Occupied File for ASA Challenge_CSV.csv"
csv08 <- "/Users/malcolm_mashig/Downloads/NYCHVS 2008 Occupied File for ASA Challenge_CSV.csv"
csv11 <- "/Users/malcolm_mashig/Downloads/NYCHVS 2011 Occupied File for ASA Challenge_CSV.csv"
csv14 <- "/Users/malcolm_mashig/Downloads/NYCHVS 2014 Occupied File for ASA Challenge_CSV.csv"
csv17 <- "/Users/malcolm_mashig/Downloads/NYCHVS 2017 Occupied File for ASA Challenge_CSV.csv"

relevant_cols <- c('borough' = 'Borough', 'building_condition' = 'Condition of building', 'age' = "Householder's Age Recode", 'sex' = "Householder's Sex", 'race' = "Householder's Race", 'hispanic_origin' = "Householder's Hispanic Origin", "move_in_year" = 'Year Householder Moved into Unit', 'mortgage' = "Mortgage Status", 'num_of_rooms' = "Number of rooms", 'num_of_bedrooms' = "Number of bedrooms", 'rent' = "Monthly contract rent", 'num_of_persons' = "Number of Persons Recode", 'gross_rent' = "Monthly Gross Rent", 'total_income' = "Total Household Income Recode") ## 15 total

data91 <- read_csv(csv91, skip = 1) %>% 
  as_tibble() %>% 
  select(relevant_cols) %>% 
  mutate('data_year' = 1991)

data93 <- read_csv(csv93, skip = 1) %>% 
  as_tibble() %>% 
  select(relevant_cols) %>% 
  mutate('data_year' = 1993)

data96 <- read_csv(csv96, skip = 1) %>% 
  as_tibble() %>% 
  select(relevant_cols) %>% 
  mutate('data_year' = 1996)

data99 <- read_csv(csv99, skip = 1) %>% 
  as_tibble() %>% 
  select(relevant_cols) %>% 
  mutate('data_year' = 1999)

data02 <- read_csv(csv02, skip = 1) %>% 
  as_tibble() %>% 
  select(relevant_cols) %>% 
  mutate('data_year' = 2002)

data05 <- read_csv(csv05, skip = 1) %>% 
  as_tibble() %>% 
  select(relevant_cols) %>% 
  mutate('data_year' = 2005)

data08 <- read_csv(csv08, skip = 1) %>% 
  as_tibble() %>% 
  select(relevant_cols) %>% 
  mutate('data_year' = 2008)

data11 <- read_csv(csv11, skip = 1) %>% 
  as_tibble() %>% 
  select(relevant_cols) %>% 
  mutate('data_year' = 2011)

data14 <- read_csv(csv14, skip = 1) %>% 
  as_tibble() %>% 
  select(relevant_cols) %>% 
  mutate('data_year' = 2014)

data17 <- read_csv(csv17, skip = 1) %>% 
  as_tibble() %>% 
  select(relevant_cols) %>% 
  mutate('data_year' = 2017)

merge1 <- full_join(data91, data93)
merge2 <- full_join(data96, data99)
merge3 <- full_join(data02, data05)
merge4 <- full_join(data08, data11)
merge5 <- full_join(data14, data17)
merge6 <- full_join(merge1, merge2)
merge7 <- full_join(merge3, merge4)
merge8 <- full_join(merge6, merge7)
merged_data <- full_join(merge8, merge5)

# check to see if all the data is represented after merge

count(data91) + count(data93) + count(data96) + count(data99) + count(data02) + count(data05) + count(data08) + count(data11) + count(data14) + count(data17) == count(merged_data)

# Decoding ----------------------------------------------------------------

# borough

merged_data[merged_data$borough == 1, 'borough'] <- 'bronx'
merged_data[merged_data$borough == 2, 'borough'] <- 'brooklyn'
merged_data[merged_data$borough == 3, 'borough'] <- 'manhattan'
merged_data[merged_data$borough == 4, 'borough'] <- 'queens'
merged_data[merged_data$borough == 5, 'borough'] <- 'staten_island'

# building condition

merged_data[merged_data$building_condition == 1, 'building_condition'] <- 'dilapidated'
merged_data[merged_data$building_condition == 2, 'building_condition'] <- 'sound'
merged_data[merged_data$building_condition == 3, 'building_condition'] <- 'detoriating'
merged_data[merged_data$building_condition == 8, 'building_condition'] <- NA

# Sex

merged_data[merged_data$sex == 1, 'sex'] <- 'male'
merged_data[merged_data$sex == 2, 'sex'] <- 'female'
merged_data[merged_data$sex == 8, 'sex'] <- NA

# Hispanic

merged_data[merged_data$hispanic_origin == 1, 'hispanic_origin'] <- 'non-hispanic'
merged_data[merged_data$hispanic_origin == 2, 'hispanic_origin'] <- 'puerto_rican'
merged_data[merged_data$hispanic_origin == 3, 'hispanic_origin'] <- 'dominican'
merged_data[merged_data$hispanic_origin == 4, 'hispanic_origin'] <- 'cuban'
merged_data[merged_data$hispanic_origin == 5, 'hispanic_origin'] <- 'so_ce_american'
merged_data[merged_data$hispanic_origin == 6, 'hispanic_origin'] <- 'mexican_am_chic'
merged_data[merged_data$hispanic_origin == 7, 'hispanic_origin'] <- 'other_span'
merged_data[merged_data$hispanic_origin == 8, 'hispanic_origin'] <- NA

# Race

merged_data[merged_data$race == 01, 'race'] <- 'white'
merged_data[merged_data$race == 02, 'race'] <- 'black'
merged_data[merged_data$race == 03, 'race'] <- 'native'
merged_data[merged_data$race == 04, 'race'] <- 'chinese'
merged_data[merged_data$race == 05, 'race'] <- 'filipino'
merged_data[merged_data$race == 06, 'race'] <- 'korean'
merged_data[merged_data$race == 07, 'race'] <- 'vietnamese'
merged_data[merged_data$race == 08, 'race'] <- 'asian_indian'
merged_data[merged_data$race == 09, 'race'] <- 'other_asian'
merged_data[merged_data$race == 10, 'race'] <- 'other'
merged_data[merged_data$race == 98, 'race'] <- NA

# Mortgage

merged_data[merged_data$mortgage == 1, 'mortgage'] <- 'yes'
merged_data[merged_data$mortgage == 2, 'mortgage'] <- 'owner'
merged_data[merged_data$mortgage == 8, 'mortgage'] <- NA

# Number of rooms

merged_data[merged_data$num_of_rooms == 9, 'num_of_rooms'] <- NA

# rent amount
merged_data[merged_data$rent == 99998, 'rent'] <- NA

# gross rent

merged_data[merged_data$gross_rent == 99998, 'gross_rent'] <- NA
merged_data[merged_data$gross_rent == 9999, 'gross_rent'] <- NA

# income

merged_data[merged_data$total_income == 999998, 'total_income'] <- NA
merged_data[merged_data == 999999] <- NA
merged_data[merged_data == 9999999] <- NA

# number of bedrooms

merged_data[merged_data$num_of_bedrooms == 98, 'num_of_bedrooms'] <- NA

# in general

merged_data[merged_data == 99999] <- NA
merged_data[merged_data == 9998] <- NA

```

## Major Findings

#### **1. Over time, renters have sacrificed non-essential rooms**

Associated Code:

**NOTE** The variable `extra_rooms` is defined as the number of rooms that are not used bedrooms. That said, it is calculated as the difference between the number of rooms and number of bedrooms or between the number of rooms and number of persons (when there are more bedrooms than people and therefore bedrooms that are not being used).

```{r}
merged_data1 <- merged_data %>% 
  filter(num_of_persons <= num_of_bedrooms) %>% 
  mutate(extra_rooms = num_of_rooms - num_of_persons)

merged_data2 <- merged_data %>% 
  filter(num_of_persons > num_of_bedrooms) %>% 
  mutate(extra_rooms = num_of_rooms - num_of_bedrooms)

merged_data <- full_join(merged_data1, merged_data2)

renters <- merged_data %>% 
  filter(mortgage == 9)

extra_room_by_year <- renters %>% 
  group_by(data_year) %>% 
  summarise(mean_extra_room = mean(extra_rooms, na.rm = TRUE))

extra_room_by_year %>% as.data.frame()

ggplot(extra_room_by_year, aes(data_year, mean_extra_room)) + 
  labs(x = "Year", y = "Non-essential Rooms", title = "Average number of non-essential rooms for NYC renters over time") +
  geom_point() + 
  geom_smooth(se = FALSE)

```

This results from the fact that, over time, the average number of rooms per person has generally decreased while the average number of bedrooms per person has generally increased.

```{r}
rooms_per_person <- renters %>% 
  mutate(room_per_person = num_of_rooms / num_of_persons, 
         bedroom_per_person = num_of_bedrooms / num_of_persons) %>% 
  group_by(data_year) %>% 
  summarise(mean_room_per_person = mean(room_per_person, na.rm = TRUE),
            mean_bedroom_per_person = mean(bedroom_per_person, na.rm = TRUE))

rooms_per_person %>% as.data.frame()
  
```

This decrease in non-essential rooms for NYC renters is not exactly consistent among all boroughs. The decrease is greatest in Staten Island and Queens.

```{r}
extra_room_by_year.bor <- renters %>% 
  group_by(data_year, borough) %>% 
  summarise(mean_extra_room = mean(extra_rooms, na.rm = TRUE))

ggplot(extra_room_by_year.bor, aes(data_year, mean_extra_room, color = borough)) + 
  labs(x = "Year", y = "Non-essential Rooms", title = "Average number of non-essential rooms for NYC renters over time (by borough)") +
  geom_point() + 
  geom_smooth(se = FALSE)
```



